Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2018-12-31 man/ms.Rmd
update figure numbers

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

@@

## Species and study site

@@

During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3). We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015 (see Appendix S3).

@@

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Appendix S4).

@@

The trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2). `REVIEW meter en Discussion` This can be likely explained as a consequence of their different management origin (*i.e.* land-use legacy), as shown by the release event in SJ suggesting their coppice origin and (references). `END REVIEW` Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4). `REVIEW` In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations. `END REVIEW` For southern population differences in growth were observed. CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figure 4; Appendix S7).

For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low). Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure 5).

@@

The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8). Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed some weak suppression episodes, but not in the last 50 years.

The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8). Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed no release event except one for CA-High at the beginning of the `REVIEW`1840s? and no suppression event in the last 50 years.

@@

*Q. pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 3). The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2). Resistance values to drought for greenness and tree-growth varied between sites (Table 3). Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2). It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).

Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Tables 3). In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1). An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 3, Table S2). Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2). For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 3; Table S1). Recovery values for tree-growth of southern populations were below or close to 1 (Figure 3, Table S2).

Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 3). Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3). Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2). For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 3). For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).

The strength of the response of the resilience indices was related to drought severity. The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data `REVIEW`(1900 - 2017), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S5). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155). Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S5) even not being the most severe drought events.

@@

## Climate and tree-growth
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 6a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 6b), specially for the northern population (r > 0.6). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 6c and 6d). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees.

@@

Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction in greenness and especially in the growth of *Q. pyrenaica*. During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q. pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013]. 
Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure 5). The decline in growth observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].

## Oaks show high resilience in response to recent drought events
Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Appendix S3), we found a positive trend for vegetation greenness of the forests of *Q. pyrenaica* located in their rear edge. Our results agree with those obtained by previous works using other remote-sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species.

@@

Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 3). La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Appendix S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q. pyrenaica*. Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía (long-term perspective), observamos como estas poblaciones presentan altos valores de resiliencia. Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).

Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q. pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Appendix S5). Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005]. Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas). Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q. pyrenaica* situadas en su rear-edge.

@@

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). Las poblaciones del norte, que presentan condiciones mas secas (Tabla 1), se vieron mas afectadas por los eventos de sequía. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 6), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 6a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]

>>>>

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../templates/template_ms.docx
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

@@

## Tree species and study site

@@

During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015 (see Table S3).

@@

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Figure 3).

@@

The trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2). `REVIEW meter en Discussion` This can be likely explained as a consequence of their different management origin (*i.e.* land-use legacy), as shown by the release event in SJ suggesting their coppice origin and (references). `END REVIEW` Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4). `REVIEW` In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations. `END REVIEW` For southern population differences in growth were observed. CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figures 4 and S6).

For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low). Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure S3).

@@

The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4). Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed no release event except one for CA-High at the beginning of the `REVIEW`1840s? and no suppression event in the last 50 years.

@@

The strength of the response of the resilience indices was related to drought severity. The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data `REVIEW`(1900 - 2017), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Figure 6). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155). Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Figure 6) even not being the most severe drought events. `REVIEW` (CONEXION con el siguiente parrafo)

*Q. pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 7). The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2). Resistance values to drought for greenness and tree-growth varied between sites (Table 3). Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2). It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 7).

Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Table 3). In the 3-year period after the 2005 drought, greenness was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1). An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 7, Table S2). Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2). For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 7; Table S1). Recovery values for tree-growth of southern populations were below or close to 1 (Figure 7, Table S2).

Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 7). Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3). Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2). For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 7). For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).

@@

## Climate and tree-growth
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), specially for the northern population (r > 0.6). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figure 8b and 8c). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees.

@@

Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction in greenness and especially in the growth of *Q. pyrenaica*. During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q. pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013]. 
Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure S3). The decline in growth observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].

## Oaks show high resilience in response to recent drought events
Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of the forests of *Q. pyrenaica* located in their rear edge. Our results agree with those obtained by previous works using other remote-sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species.

@@

Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 7). La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Table S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q. pyrenaica*. Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía (long-term perspective), observamos como estas poblaciones presentan altos valores de resiliencia. Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Figure 6).

Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q. pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Figure 6). Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005]. Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas). Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q. pyrenaica* situadas en su rear-edge.

@@

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). Las poblaciones del norte, que presentan condiciones mas secas (Table 1), se vieron mas afectadas por los eventos de sequía. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]
